Order is half of life -- Establishing Totally Ordered HTN Planning Problems is Half the Job (to Getting them Solved More Efficiently)

### 2022-12-15 paper/paper-mainfile.tex
some changes

@@

\section{Introduction}
Hierarchical Task Network (HTN) planning is a hierarchical approach to planning. As defined by \cite{HTNSurvey},  %  and \cite{IntroGhallab},
tasks in HTN planning are either primitive, corresponding to an action that can be taken, or compound. HTN problems have a set of methods that specify how one might achieve a given compound task, by decomposing it into a set of sub-tasks. A compound task may even decompose into itself, either directly via a method, or indirectly via a sequence of method applications. If decomposition leads to a sequence of primitive tasks executable from the initial state, then this sequence of actions is a solution to the problem, also known as a \enquote{plan}.

@@

\section{Hierarchical Planning Formalism}
Hierarchical task network planning, also known as \textbf{HTN planning}, is one form of hierarchical planning. There are many formalisations for HTN planning. The following one borrows heavily from \cite{HTNSurvey} and \cite{Geier2011TIHTNDecidability}. HTN planning can be partially ordered, also known as POHTN planning, or totally ordered, otherwise known as TOHTN planning. TOHTN planning is a special case of POHTN planning.  % except with the initial task network replaced by a initial compound task.

A \textbf{Partially Ordered Hierarchical Planning}, also called POHTN, problem P = $(D, S_I, T_I)$
is defined over some domain $D$, 
has an initial state $S_I \in 2^F$, and 
has a initial compound task $T_I$. 
The closed world assumption also holds for HTN planning.

The domain is defined as \textbf{D} = $(F, T_P, T_C, \delta, M)$.
$F$ is the finite set of facts or state variables, $T_P$ is the finite set of all possible primitive task names, $T_C$ is the finite set of all possible compound task names, and
$\delta$ is a mapping from primitive task name to action. Actions in POHTN domain also have preconditions, adds, and delete effects.
$M$ is the finite set of decomposition methods. Each one maps a compound task name to a task network. If $m \in M$, then $m=(c, tn)$, where $c \in T_C$.

A task network \textbf{tn} = $(T, \singlePrec, \alpha)$ consists of
T, which is a finite set of task identifiers (ids); 
$\singlePrec$, which is a partial order over T; and
$\alpha$, which maps task ids $\in$ T to task names in $T_C$ and $T_P$.

% replacing t  (i.e. $tn_1 \rightarrow_{t,m} tn_2$)
A method $m = (c, tn_m)$ decomposes a task network $tn_1 = (T_1, \singlePrec_1, \alpha_1)$ into
a new task network $tn_2$ by replacing $t$, if and only if $t \in T_1$, $\alpha_1(t) = c$, and $\exists tn' = (T', \singlePrec' \alpha')$ with $tn' \cong tn_m$ and $T' \cap T = \emptyset$ and
\begin{align}
tn_2 :=     &((T_1 \setminus \{t\}) \cup T',    \singlePrec_1 \cup \singlePrec' \cup \singlePrec_X,        \alpha_1 \cup \alpha') \\
\singlePrec_X :=  &\{(t_1, t_2) \in T_1 \times T'  \mid  (t_1,t) \in \singlePrec_1 \} \cup \\
&\{(t_1, t_2) \in T' \times T_1 \mid (t, t_2) \in \singlePrec_1 \}  
\end{align}
In other words, the decomposition of a compound task results in it being removed from the task network and replaced by a copy of the method’s task network. The ordering constraints
on the removed task are inherited by its replacement tasks, as defined by $\singlePrec_X$.

A \textbf{solution} to an HTN problem is a task network $tn = (T, \prec, \alpha)$ if and only if
tn can be reached via decomposing $tn_I$, all tasks are primitive, ($\forall t \in T: \alpha(t) \in T_P$), and there exists a sequence $\langle t_1, t_2 ... t_n \rangle$ of the task ids in $T$ that agrees with $\prec$ such that the application of that sequence $\langle \alpha(t_1), \alpha(t_2) ... \alpha(t_n) \rangle$ in $S_I$ is executable.

%In other words, there is no goal state for hierarchical planning -- the goal is to find an decomposition of the initial task, then any executable refinement of the resulting decomposition. For any given hierarchical problem, it is trivial to enforce a goal state is met, by replacing the real initial task with a task whose only decomposition is to the real initial task and a primitive task ordered after it whose preconditions are equivalent to any \enquote{goal state} desired. Thus the fact that a goal state is not strictly required is not a restriction to the expressiveness of hierarchical planning. Whereas in classical planning, one only finds any executable sequence of actions to achieve a goal state, so HTN planning poses additional restrictions on which action sequences may be considered.

\textbf{Totally Ordered Hierarchical Planning}, also called TOHTN planning, is the same as partially ordered planning in all respects except the kind of task networks it allows.
%For both planning formalisms, a method $m$ maps a task $t$ to a task network \textbf{tn} = $(T, \prec, \alpha)$. TOHTN planning domains require that $\prec$ must specify a total order between task ids in $T$. This leads to a difference in expressiveness and decideability of TOHTN vs POHTN planning. POHTN planning is more expressive in general (both in terms of plan existence and in terms of computational complexity).

% As per \cite{LanguageClassificationPlanning}, if regarded from the standpoint of formal grammars, TOHTN planning is exactly as expressive as context free languages, whereas POHTN planning is strictly more expressive than context-free languages, and strictly less expressive than context-sensitive languages.
In terms of complexity classes, \cite{ErolHTNExpressivity} proved that POHTN planning is semi-decidable, whereas \cite{Alford2015TightHTNBounds} proved that, assuming arbitrary recursion, TOHTN planning is 2-EXPTIME-complete with variables, and EXPTIME-complete without.

\section{Inference of Preconditions and Effects}
The basic linearization process consists of inferring preconditions and effects for compound tasks, inferring additional orderings to the sub-tasks based on the preconditions and effects (inferred or \enquote{real} in the case of primitive tasks) of method sub-tasks. We first attempt to add them all, and then randomly remove orderings that conflict with the originally required orderings, or each other. This random removal of contradicting orderings is called \textbf{\textit{cycle-breaking}}.

\subsection{Simple Inference of Preconditions and Effects for Compound Tasks (For Lifted and Grounded Domain) }
In a lifted format, tasks and methods accept parameters, referring to any object of a specific type. Thus actions within a method have effects and preconditions that are symmetrical in terms of interactions within the method.

E.g.\ the task \textit{navigate(waypoint1, waypoint2)} will end up with some set of actions  $\bar{a}$ once decomposed. If you replace the passed parameters \textit{waypoint1} with \textit{waypoint3} and \textit{waypoint2} with \textit{waypoint4}, then \textit{navigate(waypoint3, waypoint4)} will have the same set of actions $\bar{a}$. This is true for all possible tasks by simple induction over the decomposition process.

Thus for each method, we can \enquote{ground} it in the standard way, but using an arbitrary object instead of all objects that exist in the domain.

Specifically, let $\sigma$ be the set of parameters for the method. Then for each parameter of the method, the set of typed objects $\omega$ has one object of the corresponding type, and $|\omega| = |\sigma|$. Then $\sigma(a, \omega)$, the groundings of a transition schema $a$ over $\omega$, corresponds to the ground method obtained by substituting $\sigma$ with the corresponding object of correct type taken from $\omega$.

Where a task accepts two parameters of the same type under different names, it is assumed those are different objects. For example, for \textit{Navigate(pt0, pt1, rover)}, \textit{pt0} and \textit{pt1} are assumed as different objects. % This is illustrated in Figure~\ref{CompoundPreEffForLifted}.

This allows actions to have some ground \enquote{predicate} \textit{at\_waypoint1}, rather than $at(?waypoint)$. The resulting \enquote{predicates} can be used to order the sub-tasks, while groundings for methods are discarded. Only one linearized lifted method per partially-ordered lifted method is produced.

For a grounded method, there are no parameters in a task or method. Methods decompose to specific actions, actions have preconditions and effects on specific state variables. We can then use these to produce a linearized grounded method per grounded method.

In summary, for a lifted domain, where a predicate $f$ accepts parameters $p_1, ..., p_n$, 
$f' = f(o_1, ..., o_n)$, where $o_1, ..., o_n$ are arbitrary objects of the appropriate type. For a grounded model, $f' = f$. The inferred preconditions and effects for a task are then defined as below:
\begin{align*}
& \forall t \in T_P : \PreS(t) := \Pre(t) \\
& \forall t \in T_P : \AddS(t) := \Add(t) \\
& \forall t \in T_P : \DelS(t) := \Del(t)  \\ %  \land
& \forall t \in T_C : \PreS(t) := \{f'  \mid  \exists m \in M  \\
& : m=(t,(T, \singlePrec, \alpha)) \forall t' \in T.  \forall  f \in \PreS(t') \}   \\
& \forall t \in T_C : \AddS(t) := \{f'  \mid  \exists m \in M : \\
& m=(t,(T, \singlePrec, \alpha)) \forall t' \in T.  \forall  f \in \AddS(t') \}   \\
& \forall t \in T_C : \DelS(t) := \{f'  \mid  \exists m \in M : \\
& m=(t,(T, \singlePrec, \alpha)) \forall t' \in T.  \forall  f \in \DelS(t') \}   \\ 
\end{align*}

\subsection{Complex Inference of Preconditions and Effects for Compound Tasks (for Grounded Domain) }
The work by~\cite{ConnyPreEstimation}, details a way to infer executability-relaxed preconditions and effects. Specifically, for a compound task $c$, they are a subset of the mentioned literals described in the Simple Inference section, and a super-set of the actual preconditions and effects for any given refinement and execution of a task $c$.

@@

\begin{algorithm}[h]
	\caption{Calculation of linearized methods}
	\label{alg:Algorithm1}
	\KwData{$(F, T_P, T_C, \delta, M)$}
	\KwResult{$(F, T_P, T_C, \delta, M)$}

	\For {$m=(t, (T_m, \singlePrec, \alpha)) \in M$} 
	{
		\tcc{An edge $(t, t')$ in $G$ means $t$ is ordered before $t'$}  
		$G \gets \singlePrec$ \\
		\label{alg:StartBuildGraph}
		\For{a $\in$ F}{  
			\For{$t \in T_m$}{
				\For{$t' \in T_m$}{
					if $a \in \AddS(t)$ and $a \in \PreS(t')$, add $(t, t')$ to $G$  \\ 
					if $a \in \AddS(t)$ and $a \in \DelS(t')$, add $(t', t)$ to $G$  \\ 
					if $a \in \DelS(t)$ and $a \in \PreS(t')$, add $(t', t)$ to $G$  \\ 
					if $a \in \DelS(t)$ and $a \in \AddS(t')$, add $(t, t')$ to $G$ \\  
				}
			} 
			\label{alg:EndBuildGraph}	
		}	
		\label{alg:StartCycleBreak}
		\While{$G$ has cycles in it}{
			Delete a random ordering in $G$ that is not in $\singlePrec$ \\   % $\PreS, \DelS, \AddS$, rather than $prec$ \;
		}
		\label{alg:EndCycleBreak}
		$\prec'$ = Any linearization of $G$ \\
		\label{alg:TopSort}
		$m' = (tasks(m), \prec', \alpha(t))$ \\
		$M' = M' \cup  \{m'\}$ \\

	}
	\Return $D' = (F, T_P, T_C, \delta, M')$ \\
\end{algorithm}

From these inferred preconditions and effects $\PreS(t), \AddS(t), \DelS(t)$, (which may have been derived using the ground or lifted variant), we can run Algorithm~\ref{alg:Algorithm1} to get an linearized problem.
For an visual demonstration of what this algorithm is doing:

 \begin{figure}[h]
	\scalebox{0.75}{
		\begin{subfigure}{4.5cm} 
			\begin{tikzpicture}[auto,node distance=1.2cm,
			thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries},
			square effect/.style={rectangle,draw,font=\sffamily\Large\bfseries}]

@@

\begin{figure}[h]
	\scalebox{0.75}{
		\begin{subfigure}{7cm} 
			\begin{tikzpicture}[auto,node distance=1.2cm,
			thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries},
			square effect/.style={rectangle,draw,font=\sffamily\Large\bfseries}]

@@

			\caption{Resulting method ordering has a cycle}

@@

	\scalebox{0.75}{
		\begin{subfigure}{6cm} 
			\begin{tikzpicture}[auto,node distance=1.2cm,
			thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries},
			square effect/.style={rectangle,draw,font=\sffamily\Large\bfseries}]

@@

			\caption{Delete any random ordering to break it}
		\end{subfigure}
	}  
\end{figure}

>>>>

\scheme{SinglePrecAlt}{0}{
	text = {\textbf{D}},
	pre  = {a}
}

\scheme{SingleAddAlt}{0}{
	text = {\textbf{E}},
	effs = { a}
}

@@

\section{Introduction}
\input{01--intro.tex}
Hierarchical Task Network (HTN) planning is a hierarchical approach to planning. As defined by \cite{HTNSurvey},  %  and \cite{IntroGhallab},
tasks in HTN planning are either primitive, corresponding to an action that can be taken, or compound. HTN problems have a set of methods that specify how one might achieve a given compound task, by decomposing it into a set of sub-tasks. A compound task may even decompose into itself, either directly via a method, or indirectly via a sequence of method applications. If decomposition leads to a sequence of primitive tasks executable from the initial state, then this sequence of actions is a solution to the problem, also known as a \enquote{plan}.

@@

\section{Hierarchical Planning Formalism}
\input{02--htn.tex}

\section{Linearizing Grounded HTN Problems}
\input{03--linearization-ground.tex}

% The basic linearization process consists of inferring preconditions and effects for compound tasks, inferring additional orderings to the sub-tasks based on the preconditions and effects (inferred or \enquote{real} in the case of primitive tasks) of method sub-tasks. We first attempt to add them all, and then randomly remove orderings that conflict with the originally required orderings, or each other. This random removal of contradicting orderings is called \textbf{\textit{cycle-breaking}}.

% \subsection{Simple Inference of Preconditions and Effects for Compound Tasks (For Lifted and Grounded Domain) }

% In a lifted format, tasks and methods accept parameters, referring to any object of a specific type. Thus actions within a method have effects and preconditions that are symmetrical in terms of interactions within the method.

% E.g.\ the task \textit{navigate(waypoint1, waypoint2)} will end up with some set of actions  $\bar{a}$ once decomposed. If you replace the passed parameters \textit{waypoint1} with \textit{waypoint3} and \textit{waypoint2} with \textit{waypoint4}, then \textit{navigate(waypoint3, waypoint4)} will have the same set of actions $\bar{a}$. This is true for all possible tasks by simple induction over the decomposition process.

% Thus for each method, we can \enquote{ground} it in the standard way, but using an arbitrary object instead of all objects that exist in the domain.

% Specifically, let $\sigma$ be the set of parameters for the method. Then for each parameter of the method, the set of typed objects $\omega$ has one object of the corresponding type, and $|\omega| = |\sigma|$. Then $\sigma(a, \omega)$, the groundings of a transition schema $a$ over $\omega$, corresponds to the ground method obtained by substituting $\sigma$ with the corresponding object of correct type taken from $\omega$.

% Where a task accepts two parameters of the same type under different names, it is assumed those are different objects. For example, for \textit{Navigate(pt0, pt1, rover)}, \textit{pt0} and \textit{pt1} are assumed as different objects. % This is illustrated in Figure~\ref{CompoundPreEffForLifted}.

% This allows actions to have some ground \enquote{predicate} \textit{at\_waypoint1}, rather than $at(?waypoint)$. The resulting \enquote{predicates} can be used to order the sub-tasks, while groundings for methods are discarded. Only one linearized lifted method per partially-ordered lifted method is produced.

% For a grounded method, there are no parameters in a task or method. Methods decompose to specific actions, actions have preconditions and effects on specific state variables. We can then use these to produce a linearized grounded method per grounded method.

% In summary, for a lifted domain, where a predicate $f$ accepts parameters $p_1, ..., p_n$, 
% $f' = f(o_1, ..., o_n)$, where $o_1, ..., o_n$ are arbitrary objects of the appropriate type. For a grounded model, $f' = f$. The inferred preconditions and effects for a task are then defined as below:
% \begin{align*}
% & \forall t \in T_P : \PreS(t) := \Pre(t) \\
% & \forall t \in T_P : \AddS(t) := \Add(t) \\
% & \forall t \in T_P : \DelS(t) := \Del(t)  \\ %  \land
% & \forall t \in T_C : \PreS(t) := \{f'  \mid  \exists m \in M  \\
% & : m=(t,(T, \singlePrec, \alpha)) \forall t' \in T.  \forall  f \in \PreS(t') \}   \\
% & \forall t \in T_C : \AddS(t) := \{f'  \mid  \exists m \in M : \\
% & m=(t,(T, \singlePrec, \alpha)) \forall t' \in T.  \forall  f \in \AddS(t') \}   \\
% & \forall t \in T_C : \DelS(t) := \{f'  \mid  \exists m \in M : \\
% & m=(t,(T, \singlePrec, \alpha)) \forall t' \in T.  \forall  f \in \DelS(t') \}   \\ 
% \end{align*}

% \subsection{Complex Inference of Preconditions and Effects for Compound Tasks (for Grounded Domain) }
% The work by~\cite{ConnyPreEstimation}, details a way to infer executability-relaxed preconditions and effects. Specifically, for a compound task $c$, they are a subset of the mentioned literals described in the Simple Inference section, and a super-set of the actual preconditions and effects for any given refinement and execution of a task $c$.

@@

% \begin{algorithm}[h]
% 	\caption{Calculation of linearized methods}
% 	\label{alg:Algorithm1}
% 	\KwData{$(F, T_P, T_C, \delta, M)$}
% 	\KwResult{$(F, T_P, T_C, \delta, M)$}

% 	\For {$m=(t, (T_m, \singlePrec, \alpha)) \in M$} 
% 	{
% 		\tcc{An edge $(t, t')$ in $G$ means $t$ is ordered before $t'$}  
% 		$G \gets \singlePrec$ \\
% 		\label{alg:StartBuildGraph}
% 		\For{a $\in$ F}{  
% 			\For{$t \in T_m$}{
% 				\For{$t' \in T_m$}{
% 					if $a \in \AddS(t)$ and $a \in \PreS(t')$, add $(t, t')$ to $G$  \\ 
% 					if $a \in \AddS(t)$ and $a \in \DelS(t')$, add $(t', t)$ to $G$  \\ 
% 					if $a \in \DelS(t)$ and $a \in \PreS(t')$, add $(t', t)$ to $G$  \\ 
% 					if $a \in \DelS(t)$ and $a \in \AddS(t')$, add $(t, t')$ to $G$ \\  
% 				}
% 			} 
% 			\label{alg:EndBuildGraph}	
% 		}	
% 		\label{alg:StartCycleBreak}
% 		\While{$G$ has cycles in it}{
% 			Delete a random ordering in $G$ that is not in $\singlePrec$ \\   % $\PreS, \DelS, \AddS$, rather than $prec$ \;
% 		}
% 		\label{alg:EndCycleBreak}
% 		$\prec'$ = Any linearization of $G$ \\
% 		\label{alg:TopSort}
% 		$m' = (tasks(m), \prec', \alpha(t))$ \\
% 		$M' = M' \cup  \{m'\}$ \\

% 	}
% 	\Return $D' = (F, T_P, T_C, \delta, M')$ \\
% \end{algorithm}

From these inferred preconditions and effects $\PreS(t), \AddS(t), \DelS(t)$, (which may have been derived using the ground or lifted variant), we can run Algorithm~\ref{alg:Algorithm1} to get an linearized problem.
For an visual demonstration of what this algorithm is doing, see Figure~\ref{fig:OrderingRules} and \ref{fig:CycleBreaking}.% and \ref{fig:NotCycleBreaking}:

 \begin{figure}[h]
 	\caption{Example of Desired Additional Orderings}
 	\label{fig:OrderingRules}
	\scalebox{0.75}{
		\begin{subfigure}{4.5cm} 
			\begin{tikzpicture}[auto,node distance=1.2cm,
			thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries},
			square effect/.style={rectangle,draw,font=\sffamily\Large\bfseries}]

@@

\begin{figure}[h]
	\caption{Example of Cycle-Breaking the Desired Additional Orderings}
	\label{fig:CycleBreaking}
	\scalebox{0.75}{
		\begin{subfigure}{7cm} 
		 	\centering
			\caption{Resulting method ordering has a cycle}
			\label{fig:CycleBreaking1}
			\begin{tikzpicture}[auto,node distance=1.2cm,
			thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries},
			square effect/.style={rectangle,draw,font=\sffamily\Large\bfseries}]

@@

@@

	\scalebox{0.75}{
		\begin{subfigure}{6cm} 
		 %	\centering
			\caption{Delete any random ordering to break it}
			\label{fig:CycleBreaking2}
			\begin{tikzpicture}[auto,node distance=1.2cm,
			thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries},
			square effect/.style={rectangle,draw,font=\sffamily\Large\bfseries}]

@@

%	\scalebox{0.75}{
%		\begin{subfigure}{7cm} 
%				\caption{Example of Not Needing to Cycle-Break the Desired Additional Orderings}
%			\label{fig:NotCycleBreaking}
%			\begin{tikzpicture}[auto,node distance=1.2cm,
%			thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries},
%			square effect/.style={rectangle,draw,font=\sffamily\Large\bfseries}]

%			\action{B}{SingleAddAlt,
%				width  = 2.5em, 
%				height = 2.5em, 
%				pre length = 1em,
%				eff length = 1em 
%			};

%			\action{A}{SinglePrecAlt,
%				width  = 2.5em, 
%				height = 2.5em, 
%				pre length = 1em,
%				eff length = 1em,
%				body = {right = 4em of first} 		
%			};

%			\path[every node/.style={font=\sffamily\small}]
%			(B) edge [->,bend right] node [left]  {} (A);
%			\end{tikzpicture}

%			\caption{Resulting method ordering has no cycle}
%			
%		\end{subfigure}
%	}   
\end{figure}
